Seismic Time Series Analysis\\Statistics for Spatio-Temporal Data\\\textit{-- project draft --}

### 2020-06-10 draft.tex
cleanup some labels
various tweaks to the report

\subsection{Data cleanup and transformation}\label{sec:cleanup-transform}
\subsubsection{Data gathering and cleanup}
The data is provided in CSV format, one file per day, with only two columns related to the min-max pair \cite{Data20200402} and no column related to time. This causes two problems:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item For merging days, we need to add a column with the second-precise datetime
	\item In days with missing values there are placeholder values instead a missing row, that would make the following rows misinterpreted for measurements on the wrong second. The placeholder values are one of these two min-max pairs: $(0,0)\, (-1,1)$
\end{enumerate}

Data from each station have different domains, as distinct equipment can register a specific amplitude of the seismic movements. In fact, we found some values\footnote{Values out of domain for both stations are less than 10 out of $\approx50$ million.} outside the domains, and we considered them as missing values.

In particular, we find that out of $\approx50$ million seconds, from 27-10-2018 to 28-05-2020, only 0.67\% (Uccle) and 0.37\% (Membach) have missing values. We will see later how they are distributed for each station.

@@

After aggregating by hour, the percentages of missing values remain quite stable. As can be seen from \figurename{s} \ref{missing-values-seconds:uccs} and \ref{missing-values-seconds:mems}, missing values at both stations are grouped in a few consecutive periods.
To deal with them when R methods need complete timeseries, we define different completion strategies, depending on the seasonality, using the R function \texttt{na.aggregate}.
%
\begin{figure}[t]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/missing values hour analysis-1.pdf}
		\caption{Uccle station}
		\label{missing-values-seconds:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/missing values hour analysis-1.pdf}
		\caption{Membach station}
		\label{missing-values-seconds:mems}
	\end{subfigure}
	\caption{Missing values representation}
	\label{missing-values-seconds}
\end{figure}
%
\subsubsection{Seasonality assumptions}
The already cited article \cite{NatureCoronavirusSeismic} suggests the existence of a weekly seasonality. Looking at the data it seems confirmed, in addition to a daily seasonality (\figurename~\ref{fig:zoom}).

Since our main focus is on the seasonalities themselves, we aim to prevent assumptions based on this objective. In other words, we want to avoid searching for correlation with human activity while imposing human-based seasonality periods. Thus, we use an analytical method to extract seasonalities that the data itself suggests to us. Still, we recognize that the already performed transformations impose an human-based time alteration to the data itself (Section \ref{sec:cleanup-transform}).
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/whole data plot-2"}
		\caption{Uccle station}
		\label{fig:zoom:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/whole data plot-2"}
		\caption{Membach station}
		\label{fig:zoom:mems}
	\end{subfigure}
	\caption{Zoom from January $6^{th}$ to February $3^{rd}$ 2019}
	\label{fig:zoom}
\end{figure}
%
The method we are going to apply is based on the Fourier Transform (FT), and generates a \textit{periodgram} highlighting the most interesting frequencies that might correspond to a seasonality component.
Applying the FT for any frequency, we transform our 1-dimensional data into a 2-dimensional data mapped around a circle. Then, we can sum all the resulting vectors (that are our new data points) to obtain a vector for each frequency, whose absolute value represents the power level of the same frequency. An high power level represents a significant periodicity on that frequency in our original data.

A common operation to increase the difference between significant and non-significant periodicities is to apply a MA smoothing filter with a smaller window than the interesting ones. We will use such a filter with $p=2$.

On this note we must specify that, when a significant periodicity is found, it influences smaller periodicities given by $f*\{2,3,\dots\}$ the significant frequency. Also, if any periodicity is found that is longer than the original dataset, it must be considered as simple chance, rather than serendipity.

Since we apply this method on a dataset indexed by seconds, the transformation of the frequency $f$ into hourly periodicity $p$ is obtained by $p=(1/f)/3600$.
We can analyse the resulting periodgrams for Uccle and Membach stations in \figurename{s} \ref{fig:periodgram:uccs} and \ref{fig:periodgram:mems}.
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/periodgram-1"}
		\caption{Uccle station}
		\label{fig:periodgram:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/periodgram-1"}
		\caption{Membach station}
		\label{fig:periodgram:mems}
	\end{subfigure}
	\caption{Periodgram results}
	\label{fig:periodgram}
\end{figure}
%
Both highlight multiple 24-hour periodicities. This is given by approximation factors, since we round the divisions result to the nearest natural number. It also highlights that using a single 24-hour seasonality could leave some information on the remaining components. Still, in both stations it is the most significant periodicity.

@@

In particular from the Uccle station, we can see a great number of spikes, that will probably undermine our models.

@@

This initial division demostrated to be computationally heavy while fitting some models with weekly seasonality, so that it would be challenging to fit more than a couple of models.
Given this issue, we choose the training/testing sets to hold 52/18 weeks: this is the greater train-to-test ratio we deem usable ($\approx 75\%/25\%$), while maintaining computational feasibility.
The assessing of the lockdown influence will be discussed in Section~\ref{sec:methodology}.

Some R commands cannot handle timeseries with missing values, but as we have seen both stations datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer index.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R terms \texttt{na.aggregate(x, floor)}.

\subsubsection{Decomposition}
In order to decompose our dataset into trend and seasonality(-ties) components, we followed these steps:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply a smoothing filter to extract the trend component
	\item Remove the trend from the dataset and extract the seasonal figure using the local mean method
	\item Remove the trend and the seasonal components to obtain the residuals
\end{enumerate}
While doing so, we tested different smoothing filters: simple filter (p=seasonality), Spencer's 15-point filter and MA smoothing filter (f=seasonality). Since they do not produce particularly different seasonalities, we choose to keep the results obtained by applying the MA smoothing filter, for its significance in terms of seasonally adjusted data.
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Uccle station}
		\label{fig:decomposed:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Membach station}
		\label{fig:decomposed:mems}
	\end{subfigure}
	\caption{Timeseries decomposition}{\centering\small Only the second method results are shown for clarity. Week 10 includes January $1^{st}$.\\}
	\label{fig:decomposed}
\end{figure}
%
Since the periodgrams suggested the presence of two significative periodicities, we also tested the extraction of both corresponding seasonalities:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply the MA smoothing filter (f=24) on the dataset to extract the trend component
	\item Remove the trend from the dataset and extract the 24-hour seasonal figure using the local mean method
	\item Remove the 24-hour seasonal components to obtain a deseasoned dataset
	\item Apply the MA smoothing filter (f=168) on the deseasoned dataset to extract the updated trend component
	\item Remove the trend from the already deseasoned dataset and extract the 168-hour seasonal figure using the local mean method
	\item Repeat steps 1-5 using the last deseasoned dataset until convergence (2-3 cycles)
	\item Removing the last fitted trend and the two seasonalities we obtain the final residuals
\end{enumerate}
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/extract both 24 and 168-2.pdf}
		\caption{Uccle station}
		\label{fig:direct-multi-seasonality:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/extract both 24 and 168-2.pdf}
		\caption{Membach station}
		\label{fig:direct-multi-seasonality:mems}
	\end{subfigure}
	\caption{Seasonality figures}{\centering\small\textit{Direct} refers to single seasonality extraction\\\centering\textit{Multi} refers to both seasonalities extracted at the same time\\}
	\label{fig:direct-multi-seasonality}
\end{figure}
%
In \figurename~\ref{fig:direct-multi-seasonality} we see both approaches applied to both stations. Those plots show interesting properties of our seasonalities, probably given by the method (local mean) and the fact that the weekly period is an exact multiple of the daily period:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Both ways of extraction result in the same daily seasonality extracted
	\item The sum of the daily and weekly seasonalities extracted together is the same as the weekly seasonality extracted directly
\end{itemize}
Given in particular the second property, we will only use the 168-hours seasonality while fitting our models, for both the stations.

@@

Since our dataset shows a relevant trend component, resulting in non-constant mean, we consider the use of the \textit{diff}\footnote{The \textit{diff} operator applies the \textit{d}-order differences. All \textit{diff} operators we apply use $d=1$.} operator to smooth the mean.
After applying the \textit{diff} operator with lag 168, in conjunction with the lag 1 \textit{diff} or not, we come up with two possible models by checking the ACF and PACF plots:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item from using both \textit{diff}s, the ACF and PACF on the long run hint to $\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$
	\item from using just the lag 168 \textit{diff}, the PACF on the first lags and the ACF/PACF combination on the long run hint to $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$
\end{itemize}
We also employ the \textit{auto.arima} stepwise methodology\footnote{\label{footnote:auto.arima}The \textit{auto.arima} stepwise methodology fits different ARIMA models within a range of orders. Based on each result's information criterion (Akaike), it changes each time one of the \textit{p}, \textit{d}, \textit{q}, \textit{P}, \textit{D}, or \textit{Q} values, depending on the configured limits. These changes minimizes the information criteria of the final model.}, locking the \textit{diff} values to 0 or 1, to get some hints on the possible models.

@@

Also checking the Mean Absolute Percentage Error (MAPE) on the training set, they actually have the lowest three values ($\mathrm{MAPE} < 6.2\%$). Nevertheless, MAPE is not a single valid criterion for selection, since a low error predicting the training set itself does not necessarily mean a good forecast power.
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-35}
		\caption{ACF and PACF plots}
		\label{fig:arima-26}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-32}
		\caption{Q-Q plot}
		\label{fig:arima-24}
	\end{subfigure}
	\caption{Plots for $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
	\label{fig:arima}
\end{figure}
%
The ACF and PACF of the residuals for all three models are very similar (\figurename~\ref{fig:arima-26}), even if the models themselves have great differences in meaning and in number of parameters: they highlight a clear seasonal ARMA process with periodicity 24. This shows how our previous conclusion that the 168 hours seasonality could include the 24 hours one is flawed. Unfortunately, the R function we are using does not permit to estimate multiple seasonalities\footnote{\textit{Arima} method from the \textit{forecast} R package, only specifying the seasonal and non-seasonal orders.}.

@@

Also, it must be noted that a January $1^{st}$ period lies in our test dataset, that is already noticed as low-valued period (\figurename~\ref{fig:decomposed}). We can notice a clear overestimation during those couple of weeks (around week 62 on \figurename~\ref{fig:arima-forecast}).
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-37}
		\caption{$\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:arima-forecast:200011}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-69}
		\caption{$\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$}
		\label{fig:arima-forecast:412011}
	\end{subfigure}
	\caption{ARIMA forecasts}{\footnotesize$\mathrm{ARIMA}(4,1,2)(0,1,1)_{168}$ behaves like $\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$ with faster increasing variablity}
	\label{fig:arima-forecast}
\end{figure}
%
We can see in \figurename~\ref{fig:arima-forecast} the forecast behaviour for our models. We can appreciate the constant span of the confidence intervals in \figurename~\ref{fig:arima-forecast:200011}, in constrast with the more complex models that have increasing intervals further away from the training set. It must be noted that the increasing intervals include negative values, that are actually impossible to obtain given the data meaning (difference between maximum and minimum measurements).

@@

Since the $\lambda$ value approximately simplifies the Box-Cox transformation formula from $(y^\lambda-1)/\lambda$ into $(y-1)/y$, and since we only have positive values, all transformed values fall into the higher $0-1$ spectrum, in particular $0.95-1$ (\figurename~\ref{fig:mems:box-cox:data}). This will probably cause some approximation errors that could rig the final results.
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/training data transformation plots-7"}
		\caption{Transformed training set}
		\label{fig:mems:box-cox:data}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/training data transformation plots-9"}
		\caption{Normal vs. transformed data distribution}
		\label{fig:mems:box-cox:qq}
	\end{subfigure}
	\caption{Results of Box-Cox transformation}
	\label{fig:mems:box-cox}
\end{figure}
%
The transformed data still shows no constant mean nor variance, even if they show less variability around the peaks with respect to the original data, confirmed by the Q-Q plot (\figurename~\ref{fig:mems:box-cox:qq}). Thus, we consider the usage of the \textit{diff} operator, at lag 1 and 168. Interestingly, the analysis of ACF and PACF plots hints to similar models to the untransformed Uccle data:
$\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$ and $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$. All ACF and PACF plots for Membach model estimation are not shown for brevity.

@@

The Box-Pierce test statistics for both models accept the independence hypothesis between residuals. The two models have similar AICc values, and similar MAPE on the training set ($\approx6\%$). Thus, we can assess the forecasts for both models in \figurename~\ref{fig:mems:arima}.
%
\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/arima-21"}
		\caption{$\mathrm{ARIMA}(0,1,1)(0,1,1)_{168}$}
		\label{fig:mems:arima:011011}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/arima-13"}
		\caption{$\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:mems:arima:200011}
	\end{subfigure}
	\caption{ARIMA forecasts}
	\label{fig:mems:arima}
\end{figure}
%
The first model catches a rapidly-increasing trend, while the latter predicts more stable values moving the most part of the increasing trend into slightly increasing variability. Visually, both could be acceptable, even if the latter seems to better behave on \textit{normal} days losing precision on high-valued ones, while the first tries to better accomodate the high-valued days losing precision on the \textit{normal} ones.

@@

As a side note, we followed the same procedure without applying the Box-Cox transformation, but the best model that could be chosen had more parameters and was less precise than the chosen model on the transformed data.

TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO 
%\subsubsection{MSTL}

@@

\subsubsection{Seasonalities divergence}
We can still compare how the seasonalities differ with respect with each original dataset and between each other, by extracting the seasonalities for each station from before and after the lockdown (March $14^{th}$).

>>>>

\subsection{Data cleanup and transformation}\label{sec:cleanup-transform}
\subsubsection{Data gathering and cleanup}
The data is provided in CSV format, one file per day, with only two columns related to the min-max pair \cite{Data20200402} and no column related to time. This causes two problems:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item For merging days, we need to add a column with the second-precise datetime
	\item In days with missing values there are placeholder values instead a missing row, that would make the following rows misinterpreted for measurements on the wrong second. The placeholder values are one of these two min-max pairs: $(0,0),\, (-1,1)$.
\end{enumerate}

Data from each station have different domains, as distinct equipment can register a specific amplitude of the seismic movements. In fact, we found some values\footnote{Values out of domain for each station are less than 10 out of $\approx50$ million.} outside the domains, and we consider them as missing values.

In particular, we find that out of $\approx50$ million seconds, from 27-10-2018 to 28-05-2020, only 0.67\% (Uccle) and 0.37\% (Membach) are missing values. We will see later how they are distributed for each station.

@@

After aggregating by hour, the percentages of missing values remain quite stable. As can be seen from \figurename{s} \ref{missing-values-seconds:uccs} and \ref{missing-values-seconds:mems}, missing values at both stations are grouped in a few consecutive periods.
To deal with them when R methods need complete timeseries, we define different completion strategies, depending on the seasonality, using the R function \texttt{na.aggregate}.

\begin{figure}[t]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/missing values hour analysis-1.pdf}
		\caption{Uccle station}
		\label{missing-values-seconds:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/missing values hour analysis-1.pdf}
		\caption{Membach station}
		\label{missing-values-seconds:mems}
	\end{subfigure}
	\caption{Missing values representation}
	\label{missing-values-seconds}
\end{figure}
%
\subsubsection{Seasonality assumptions}
The already cited article \cite{NatureCoronavirusSeismic} suggests the existence of a weekly seasonality. Looking at the data it seems confirmed, in addition to a daily seasonality (\figurename~\ref{fig:zoom}).

Since our main focus is on the seasonalities themselves, we aim to prevent assumptions based on this objective. In other words, we want to avoid searching for correlation with human activity while imposing human-based seasonality periods. Thus, we use an analytical method to extract seasonalities that the data itself suggests to us. Still, we recognize that the already performed transformations impose an human-based time alteration to the data itself (Section \ref{sec:cleanup-transform}).

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/whole data plot-2"}
		\caption{Uccle station}
		\label{fig:zoom:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/whole data plot-2"}
		\caption{Membach station}
		\label{fig:zoom:mems}
	\end{subfigure}
	\caption{Zoom from January $6^{th}$ to February $3^{rd}$ 2019}
	\label{fig:zoom}
\end{figure}
%
The method we are going to apply is based on the Fourier Transform (FT), and generates a \textit{periodgram} highlighting the frequencies that might correspond to a seasonality component \cite[chapter 11]{brockwell1991time}.
Applying the FT for any frequency, we transform our 1-dimensional data into a 2-dimensional data mapped around a circle. Then, we can sum all the resulting vectors (that are our new data points) to obtain a vector for each frequency, whose absolute value represents the power level of the same frequency. An high power level represents a significant periodicity on that frequency in our original data.

A common operation to increase the difference between significant and non-significant periodicities is to apply a Moving Average (MA) smoothing filter with a smaller window than the interesting ones. We will use such a filter with $p=2$.

On this note we must specify that, when a significant periodicity is found, it influences smaller periodicities given by $f*\{2,3,\dots\}$ the significant frequency \cite{brockwell1991time}. Also, if any periodicity is found that is longer than the original dataset, it must be considered as simple chance, rather than serendipity.

Since we apply this method on a dataset indexed by seconds, the transformation of the frequency $f$ into hourly periodicity $p$ is obtained by $p=(1/f)/3600$.
We can analyse the resulting periodgrams for Uccle and Membach stations in \figurename{s} \ref{fig:periodgram:uccs} and \ref{fig:periodgram:mems}.
An higher power value for a frequency means that the corresponding periodicity is meaningful for describing the dataset.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.uccs_files/figure-latex/periodgram-1"}
		\caption{Uccle station}
		\label{fig:periodgram:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{"project.mems_files/figure-latex/periodgram-1"}
		\caption{Membach station}
		\label{fig:periodgram:mems}
	\end{subfigure}
	\caption{Periodgram results}
	\label{fig:periodgram}
\end{figure}
%
Both highlight multiple 24-hour periodicities. This is given by approximation factors, since we round the divisions result to the nearest natural number. It also highlights that using a single 24-hour seasonality could leave some information on the remaining components. Still, in both stations it is the most significant periodicity.

@@

@@

This initial division demostrated to be computationally heavy while fitting some models with weekly seasonality, so that it would be challenging to fit more than a couple of models.
Given this issue, we choose the training/testing sets to hold 52/18 weeks: this is the greater train-to-test ratio we deem usable ($\approx 75\%/25\%$), while maintaining computational feasibility.
The assessment of the lockdown influence will be discussed in Section~\ref{sec:methodology}.

Some R commands cannot handle timeseries with missing values, but as we have seen both stations datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer index.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R, \texttt{na.aggregate(x, floor)}.

\subsubsection{Decomposition}
In order to decompose our dataset into trend and seasonality(-ties) components, we followed these steps:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply a smoothing filter to extract the trend component
	\item Remove the trend from the dataset and extract the seasonal figure using the local mean method
	\item Remove the trend and the seasonal components to obtain the residuals
\end{enumerate}
While doing so, we tested different smoothing filters: simple filter (p=seasonality), Spencer's 15-point filter and MA smoothing filter (f=seasonality). Since they do not produce particularly different seasonalities, we choose to keep the results obtained by applying the MA smoothing filter, for its significance in terms of seasonally adjusted data.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Uccle station}
		\label{fig:decomposed:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/extract both 24 and 168-1.pdf}
		\caption{Membach station}
		\label{fig:decomposed:mems}
	\end{subfigure}
	\caption{Timeseries decomposition}{\centering\small Only the second method results are shown for clarity. Week 10 includes January $1^{st}$.\\}
	\label{fig:decomposed}
\end{figure}
%
Since the periodgrams suggested the presence of two significative periodicities, we also tested the extraction of both corresponding seasonalities:
\begin{enumerate}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Apply the MA smoothing filter (f=24) on the dataset to extract the trend component
	\item Remove the trend from the dataset and extract the 24-hour seasonal figure using the local mean method
	\item Remove the 24-hour seasonal components to obtain a deseasoned dataset
	\item Apply the MA smoothing filter (f=168) on the deseasoned dataset to extract the updated trend component
	\item Remove the trend from the already deseasoned dataset and extract the 168-hour seasonal figure using the local mean method
	\item Repeat steps 1-5 using the last deseasoned dataset until convergence (2-3 cycles)
	\item Removing the last fitted trend and the two seasonalities we obtain the final residuals
\end{enumerate}

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/extract both 24 and 168-2.pdf}
		\caption{Uccle station}
		\label{fig:direct-multi-seasonality:uccs}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.mems_files/figure-latex/extract both 24 and 168-2.pdf}
		\caption{Membach station}
		\label{fig:direct-multi-seasonality:mems}
	\end{subfigure}
	\caption{Seasonality figures}{\centering\small\textit{Direct} refers to single seasonality extraction\\\centering\textit{Multi} refers to both seasonalities extracted at the same time\\}
	\label{fig:direct-multi-seasonality}
\end{figure}
%
In \figurename~\ref{fig:direct-multi-seasonality} we see both approaches applied to both stations. Those plots show interesting properties of our seasonalities, probably given by the method (local mean) and the fact that the weekly period is an exact multiple of the daily period:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item Both ways of extraction result in the same daily seasonality extracted
	\item The sum of the daily and weekly seasonalities extracted together is the same as the weekly seasonality extracted directly
\end{itemize}
Given in particular the second property, we will only use the 168-hours seasonality while fitting our models, for both the stations.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\linewidth]{"project.uccs_files/figure-latex/diffs/d168-acf-pacf-1"}
	\caption{ACF and PACF plots after \textit{diff} at lag 168}
	\label{fig:uccs:p-acf}
\end{figure}
%
Since our dataset shows a relevant trend component, resulting in non-constant mean, we consider the usage of the \textit{diff}\footnote{The \textit{diff} operator applies the \textit{d}-order differences. All \textit{diff} operators we apply use $d=1$.} operator to smooth the mean.
After applying the \textit{diff} operator with lag 168, in conjunction with the lag 1 \textit{diff} or not, we come up with two possible models by checking the ACF and PACF plots:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item from using both \textit{diff}s, the ACF and PACF on the long run hint to $\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$ (plots omitted for brevity)
	\item from using just the lag 168 \textit{diff}, the PACF on the first lags and the ACF/PACF combination on the long run hint to $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$ (\figurename~\ref{fig:uccs:p-acf})
\end{itemize}
We also employ the \textit{auto.arima} stepwise methodology\footnote{\label{footnote:auto.arima}The \textit{auto.arima} stepwise methodology fits different ARIMA models within a range of orders. Based on each result's information criterion (Akaike), it changes each time one of the \textit{p}, \textit{d}, \textit{q}, \textit{P}, \textit{D}, or \textit{Q} values, depending on the configured limits. These changes minimizes the information criteria of the final model.}, locking the \textit{diff} values to 0 or 1, to get some hints on the possible models.

@@

Also checking the Mean Absolute Percentage Error (MAPE) on the training set, they actually have the lowest three values ($\mathrm{MAPE} < 6.2\%$). Nevertheless, MAPE is not a single valid criterion for selection, since a low error predicting the training set itself does not necessarily mean a good forecast power.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-35}
		\caption{ACF and PACF plots}
		\label{fig:arima-26}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-32}
		\caption{Q-Q plot}
		\label{fig:arima-24}
	\end{subfigure}
	\caption{Plots for $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
	\label{fig:arima}
\end{figure}
%
The ACF and PACF of the residuals for all three models are very similar (\figurename~\ref{fig:arima-26}), even if the models themselves have great differences in meaning and in number of parameters: they highlight a clear seasonal ARMA process with periodicity 24. This shows how our previous conclusion that the 168 hours seasonality could include the 24 hours one is flawed. Unfortunately, the R function we are using does not permit to estimate multiple seasonalities\footnote{\textit{Arima} method from the \textit{forecast} R package, only specifying the seasonal and non-seasonal orders.}.

@@

Also, it must be noted that a January $1^{st}$ period lies in our test dataset, that is already noticed as low-valued period (\figurename~\ref{fig:decomposed}). We can notice a clear overestimation during those couple of weeks (around week 62 on \figurename~\ref{fig:arima-forecast}).

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-37}
		\caption{$\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:arima-forecast:200011}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{project.uccs_files/figure-latex/arima-69}
		\caption{$\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$}
		\label{fig:arima-forecast:412011}
	\end{subfigure}
	\caption{ARIMA forecasts}{\footnotesize$\mathrm{ARIMA}(4,1,2)(0,1,1)_{168}$ behaves like $\mathrm{ARIMA}(3,1,1)(0,1,1)_{168}$ with faster increasing variablity}
	\label{fig:arima-forecast}
\end{figure}
%
We can see in \figurename~\ref{fig:arima-forecast} the forecast behaviour for our models. We can appreciate the constant span of the confidence intervals in \figurename~\ref{fig:arima-forecast:200011}, in constrast with the more complex models that have increasing intervals further away from the training set. It must be noted that the increasing intervals include negative values, that are actually impossible to obtain given the data meaning (difference between maximum and minimum measurements).

@@

Since the $\lambda$ value approximately simplifies the Box-Cox transformation formula from $(y^\lambda-1)/\lambda$ into $(y-1)/y$, and since we only have positive values, all transformed values fall into the higher $0-1$ spectrum, in particular $0.95-1$ (\figurename~\ref{fig:mems:box-cox:data}). This will probably cause some approximation errors that could rig the final results.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/training data transformation plots-7"}
		\caption{Transformed training set}
		\label{fig:mems:box-cox:data}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/training data transformation plots-9"}
		\caption{Normal vs. transformed data distribution}
		\label{fig:mems:box-cox:qq}
	\end{subfigure}
	\caption{Results of Box-Cox transformation}
	\label{fig:mems:box-cox}
\end{figure}
%
The transformed data still shows no constant mean nor variance, even if they show less variability around the peaks with respect to the original data, confirmed by the Q-Q plot (\figurename~\ref{fig:mems:box-cox:qq}). Thus, we consider the usage of the \textit{diff} operator, at lag 1 and 168. The analysis of ACF and PACF plots hints to two models: $\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$ and $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$. All ACF and PACF plots for Membach model estimation are not shown for brevity.

@@

The Box-Pierce test statistics for both models accept the independence hypothesis between residuals. The two models have similar AICc values, and similar MAPE on the training set ($\approx6\%$). Thus, we can assess the forecasts for both models in \figurename~\ref{fig:mems:arima}.

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/arima-21"}
		\caption{$\mathrm{ARIMA}(0,1,1)(0,1,1)_{168}$}
		\label{fig:mems:arima:011011}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=1\linewidth]{"project.mems_files/figure-latex/arima-13"}
		\caption{$\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:mems:arima:200011}
	\end{subfigure}
	\caption{ARIMA forecasts}
	\label{fig:mems:arima}
\end{figure}
%
The first model catches a rapidly-increasing trend, while the latter predicts more stable values moving the most part of the increasing trend into slightly increasing variability. Visually, both could be acceptable, even if the latter seems to better behave on \textit{normal} days losing precision on high-valued ones, while the first tries to better accomodate the high-valued days losing precision on the \textit{normal} ones.

Also, it is important to notice that while the first model includes in the confidence intervals negative values, non existent in our domain, the latter directs all the variability to higher values, while maintaining a consistent lower bound. This appears to be a good feature to correctly model our data.

@@

Since checking the ACF and PACF plots for the original data could lead to a model with the same order, we fitted it for comparison. The MAPEs are higher but comparable, $7\%$ on the test set and $19\%$ on the train set, the Box-Pierce test refutes the independence between residuals, and $88\%/94\%$ of test data fits inside the $80\%$ and $95\%$ CI. Finally, it shows a variability that tends to admit lower values than the training data shows, including negative ones, while having a lower acceptance for higher values. We omit plots related to this model, since such results confirm the usage of the already selected model on the transformed data.

It is important to highlight that no model we assessed found any pattern for predicting the high-valued spikes.
%\subsubsection{MSTL}

@@

Still, we can assess how the lockdown period is forecasted for both stations, and then compare their efficency. Also, we can extract the seasonalities, using the same MA filters we already used, separately from the periods before and during the lockdown, and compare also these results.

\subsubsection{Seasonalities divergence}
We will highlight the confidence bands for an estimated error of 67\%. This interval estimation lies on the assumption that values for each single hour in the hourly/weekly seasonality is Normally distributed, a slightly different assumption that is not necessarily refuted by the already seen Q-Q plots.
